Phylogeographic inference with microtellite and SNPs

### 2019-07-01 code/rmarkdown/Supporting_information.rmd
convergence plots

---
title: "Supporting Information: Phylogeographic inference with microsatellites and SNPs"
author: |
  | Miguel Camacho-Sanchez${\ast^1}$, Guillermo Velo-Antón$^{1}$, Ana Veríssimo$^1$, Silvia Carvalho$^2$
  | $^1$ CIBIO/InBIO, Centro de Investigação em Biodiversidade e Recursos Genéticos da Universidade do Porto, Vairão, Portugal
  | Correspondence should be addressed to miguelcamachosachez@gmail.com
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  rmarkdown::pdf_document:
    toc: false
    keep_tex: true
    fig_caption: yes
    includes:
      in_header: preamble.tex
fontsize: 11pt
documentclass: article
subparagraph: true
---

>>>>

---
title: "Supporting Information: Phylogeographic inference with microsatellites and SNPs"
author: |
  | Miguel Camacho-Sanchez${\ast^1}$, Guillermo Velo-Antón$^{1}$, Ana Veríssimo$^1$, Silvia Carvalho$^2$
  | $^1$ CIBIO/InBIO, Centro de Investigação em Biodiversidade e Recursos Genéticos da Universidade do Porto, Vairão, Portugal
  | Correspondence should be addressed to miguelcamachosachez@gmail.com
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  rmarkdown::pdf_document:
    toc: false
    keep_tex: true
    fig_caption: yes
    includes:
      in_header: preamble.tex
fontsize: 11pt
documentclass: article
subparagraph: true
---
Table S1: samples
Figure S1: SNP filtering for *H. molleri* (A, C, E, G) and *P. cultripes* (B, D, F, H). En each panel we show the frequency distribution (y-axes) of the metric evaluated (x-axes), with the thresholds applied (dashed vertical lines).

Figures 2.1 - 2.11: Change of the likelihood of the data across iterations for 10 replicate runs for K = 1 to K = 10. The vertical dashed lines indicate burnin length.

Figures S3.1 - S3.2: Gelman and Rubin's convergence diagnostic for SNP (S3.1) and microtellite (S3.2) datasets.

### 2019-07-01 code/rmarkdown/article.Rmd
convergence plots

```{r, include = FALSE}
knitr::opts_knit$set(root.dir = normalizePath("../.."))
```

```{r, include = FALSE}
# restore session
gen <- readRDS("data/intermediate/raw_genotypes.RDS")
# load packages
# define helper functions
# load data

@@

The above-mentioned sharp clustering with large-scale SNP data compared to microtellites using non-parametric methods does not reflect when using model-based clustering in the popular software STRUCTURE. This software provides a powerful tool to estimate individual ancestries from genotypes data (Pritchard et al. 2000). Simulations show equivalent power of tens of microtellites and thousands of SNPs to retrieved with high confidence the correct number of cluster (Haasl and Payseur 2011). With empirical data, overall genetic structure with both datasets can vary in the number of most optimal number of clusters (best K) and proportions of individual ancestries (Bradbury et al. 2015; Malenfant et al. 2015; Elbers et al. 2016; Bohling et al. 2019) or be practically identical (Lemopoulos et al. 2019). However, the best K can be very sensitive to demographic processes and can vary substantially depending on the method to estimate it (Janes et al. 2017; Wang 2019). Other issues as uneven sampling and hierarchical population structure can also difficult the inference of the correct population structure (Puechmaille et al. 2016; Wang 2016). Neighbor joining clustering is a complementary way to overcome some limitations of model-based clustering and to recover a reliable evolutionary history of the populations under a wide variaty of scenarios (Kalinowski 2009, 2011). The reconstructions of phylogenetic trees using neighbor joining with microtellite data are often coherent with the phylogeographic structure recovered with SNPs. However, the large homoplasy undermines their power to retrieve phylogenetic relationships between distant populations (Estoup et al. 2002). So, trees from microsatellite data recurrently have little support, specially at deeper nodes, which limits the interpretations of the evolutionary history of the populations (Moore et al. 2014; Candy et al. 2015; Bruneaux et al. 2012; Bradbury et al. 2018), although when several hundreds of loci are used the correct topology is recovered correctly (Takezaki and Nei 2006). The large amount of SNPs and their genome-wide distribution secure a range mutation rates which can provide sufficient information at different evolutionary scales, from recent within species processes to interspecies phylogenies (eg. Defaveri et al. 2013; Petersen et al. 2013).

@@

# 2. Materials and Methods
## 2.1 Data collection
We used microsatellite and SNP datasets from two Iberian amphibians, *Hyla molleri* and *Pelobates cultripes*, covering most of their native range (Figure 1).
The microsatellite genotypes were obtained from already-published studies. For *Hyla molleri*, we gathered the microsatellite genotypes from *Appendix 1* in Sánchez-Montes et al. (2019). They consisted of `r adegenet::nInd(gen$usat_hyla)` samples with `r adegenet::nLoc(gen$usat_hyla)` genotyped microsatellites. For *Pelobates cultripes*, we requested the genotypes to the authors in Gutiérrez-Rodríguez et al. (2017). This dataset consisted of `r adegenet::nInd(gen$usat_pelo)` samples with `r adegenet::nLoc(gen$usat_pelo)` genotyped microsatellites.
The SNP data was generated using DArTseq technology (Diversity Arrays Technology, Canberra, Australia). We included samples spanning all over the Iberian Peninsula (Figure 1; Table 1). DNA was extracted ExtractMe Genomic DNA 96-Well (DNA GDAŃSK), concentrated with QIAamp DNA Micro (QIAGEN GmbtH) kits, if necessary, and sent for sequencing.
##

## Study area and species

## Data

# 3 | RESULTS

* result 1: sequencing
We sequenced a total of ´r adegenet::nInd(gen)´ samples which included 1 replica. DArTseq genotyped ´r adegenet::nLoc(gen)´SNPs in ´r nlevels(as.factor(gen@other$loc.metrics$CloneID))´ loci. The overall missing data was ´r ´ ()

* result 2: structure vs dapc

>>>>

```{r, include = FALSE}
knitr::opts_knit$set(root.dir = normalizePath("../.."))
library(dplyr)
```

```{r, include = FALSE}
# data
gen <- readRDS("data/intermediate/gen_consolidated_filtered.rds")
samples <- readRDS("data/intermediate/table1.rds")

#Results
#number of individuals H. molleri
indH <- names(gen) %>% grep(pattern = "hyla") %>%
            sapply(function(x) adegenet::nInd(gen[[x]])) %>% sum
#number of individuals P. cultripes
indP <- names(gen) %>% grep(pattern = "pelo") %>%
            sapply(function(x) adegenet::nInd(gen[[x]])) %>% sum
indHm <- adegenet::nInd(gen$usat_hyla)
indHs <- adegenet::nInd(gen$dart_hyla)
indPm <- adegenet::nInd(gen$usat_pelo)
indPs <- adegenet::nInd(gen$dart_pelo)
#total individuals
totalind <- indP + indH
#localities for H. molleri
popH <- samples$samples %>% grep(pattern = "H.m.") %>% length
#localities for P. cultripes
popP <- samples$samples %>% grep(pattern = "P.c.") %>% length
#missing data microtellite
# H. molleri
missHm <- hierfstat::genind2hierfstat(gen$usat_hyla) %>%
            dplyr::select(-1) %>% {xx <<- length(as.matrix(.)); . } %>%
            is.na() %>% {sum(.) / xx * 100} %>% round(1)
# P. cultripes
missPm <- hierfstat::genind2hierfstat(gen$usat_pelo) %>%
            dplyr::select(-1) %>% {xx <<- length(as.matrix(.)); . } %>%
            is.na() %>% {sum(.) / xx * 100} %>% round(1)
#missing data SNPs
missHs <- as.matrix(gen$dart_hyla) %>%
            {xx <<- length(.); . } %>%
            is.na() %>% {sum(.) / xx * 100} %>% round(1)
missPs <- as.matrix(gen$dart_pelo) %>%
            {xx <<- length(.); . } %>%
            is.na() %>% {sum(.) / xx * 100} %>% round(1)
#Number of loci
# H. molleri
#  microtellites
locHm <- adegenet::nLoc(gen$usat_hyla)
#  SNPs
locHs <- adegenet::nLoc(gen$dart_hyla)
# P. cultripes
#  microtellites
locPm <- adegenet::nLoc(gen$usat_pelo)
#  SNPs
locPs <- adegenet::nLoc(gen$dart_pelo)

#raw raw_genotypes
genraw <- readRDS("data/intermediate/raw_genotypes.rds")
rawHm <- adegenet::nInd(genraw$usat_hyla)
rawHs <- adegenet::nInd(genraw$dart_hyla)
rawPm <- adegenet::nInd(genraw$usat_pelo)
rawPs <- adegenet::nInd(genraw$dart_pelo)

#release memory
rm(genraw, gen)

@@

The above-mentioned sharp clustering with large-scale SNP data compared to microtellites using non-parametric methods does not reflect when using model-based clustering in the popular software STRUCTURE. This software provides a powerful tool to estimate individual ancestries from genotype data (Pritchard et al. 2000). The modeling of allele frequencies can be more powerful (Nassir et al. 2008), but comes at a high computer cost compared to non-parametric clustering (Porras-Hurtado et al. 2013), and in many cases, methods such as discriminant analysis of principal components claim to perform better (Jombart et al. 2008). Simulations show equivalent power of tens of microtellites and thousands of SNPs to retrieved with high confidence the correct number of cluster (Haasl and Payseur 2011). With empirical data, overall genetic structure with both datasets can vary in the number of most optimal number of clusters (best K) and proportions of individual ancestries (Bradbury et al. 2015; Malenfant et al. 2015; Elbers et al. 2016; Bohling et al. 2019) or be practically identical (Lemopoulos et al. 2019). However, the best K can be very sensitive to demographic processes and can vary substantially depending on the method to estimate it (Janes et al. 2017; Wang 2019). Other issues as uneven sampling and hierarchical population structure can also difficult the inference of the correct population structure (Puechmaille et al. 2016; Wang 2016). Neighbor joining clustering is a complementary way to overcome some limitations of model-based clustering and to recover a reliable evolutionary history of the populations under a wide variaty of scenarios (Kalinowski 2009, 2011). The reconstructions of phylogenetic trees using neighbor joining with microtellite data are often coherent with the phylogeographic structure recovered with SNPs. However, the large homoplasy undermines their power to retrieve phylogenetic relationships between distant populations (Estoup et al. 2002). So, trees from microsatellite data recurrently have little support, specially at deeper nodes, which limits the interpretations of the evolutionary history of the populations (Moore et al. 2014; Candy et al. 2015; Bruneaux et al. 2012; Bradbury et al. 2018), although when several hundreds of loci are used the correct topology is recovered correctly (Takezaki and Nei 2006). The large amount of SNPs and their genome-wide distribution secure a range mutation rates which can provide sufficient information at different evolutionary scales, from recent within species processes to interspecies phylogenies (eg. Defaveri et al. 2013; Petersen et al. 2013).

@@

# 2. Materials and Methods
## 2.1 Data collection
We used microsatellite and SNP datasets from two Iberian amphibians, *Hyla molleri* and *Pelobates cultripes*, covering most of their native range (Figure 1; Table S1).
The microsatellite genotypes were obtained from already-published studies. For *Hyla molleri*, we gathered them from *Appendix 1* in Sánchez-Montes et al. (2019). They consisted of `r rawHm` samples with `r locHm` loci. For *Pelobates cultripes*, we requested the genotypes to the authors in Gutiérrez-Rodríguez et al. (2017). This dataset consisted of `r rawPm` samples with `r locPm` loci.
SNP data was generated for `r rawHs` samples of *H. molleri* and `r rawPs` samples of *P. cultripes*. *???for Guillermo: I pasted the following from Gutiérrez-Rogríguez et al. (2017). Please, check: Most samples were obtained from tail tips of larvae and toes of adults, pre- served in absolute ethanol, and subsequently kept in a freezer at 20 °C???*. For *H. molleri*, we completed the sampling with tissue from several museum specimens (Table S1). Genomic DNA was extracted ExtractMe Genomic DNA 96-Well (DNA GDAŃSK), concentrated with QIAamp DNA Micro (QIAGEN GmbtH) kits, if necessary, and sequenced with DArTseq technology (Diversity Arrays Technology, Canberra, Australia).

## 2.2 Data generation
### 2.2.1 SNPs sequencing

Amphibians have large genomes with repetitive regions which hampers genomic sequencing approaches (Funk et al. 2018; Liedtke et al. 2018). We chose DArTseq, a proprietary protocol to sequence reduced representation of the genome from double-digested fragments. DArTseq has been claimed to work well with amphibians because it enriches for high-complexity regions, disregarding repetitive portions of the genome (Lambert et al. 2016). DNA extracts from *H. molleri* and *P. cultripes* with total DNA ranging from to xxx ug were sent to Diversity Arrays Technology (University of Canberra, Australia). The fragments were generated using the restriction enzymes xxx and xxx, and sequenced in an Illumina xxxx as sigle-end reads of 77 nucleotides (nt). The sequencing depth for *H. molleri* and *P. cultripes* were xxx and xxx M reads / sample, respectively. Diversity Arrays Technology genotypes the samples using the proprietary DArTSoft14 pipeline and returns a csv file with the genotypes and several parameters on each SNP. A xx % of the samples in the run are included as internal replicas to provide confidence levels on the genotype calls.

### 2.2.2 Data filtering
We additionally applied several filtering steps to the genotype matrices in R 3.5.1 (R Core Team 2018), using functions from the *dartR* 1.1.11 package (Gruber and Georges 2019) and custom functions (https://github.com/csmiguel/usat_snp; Figure S1). The genotypes were imported into R and filters were applied following order. A per-sample call rate threshold (proportion loci with calls) of 0.35, RepAvg (a repeatability measure of genotypes calls between replicas) threshold of 0.95. We used thresholds for allele balance of 0.15 and 0.85, and removed loci for which their coverage was 3.5 times higher than the median coverage (O'Leary et al. 2018), after which, we filtered loci with a call rate (proportion of samples with a call) below 0.8, retained only one SNP per contig, a set minor allele frequency threshold of 0.02 and removed loci whose alleles deviated from Hardy-Weinberg proportions(O'Leary et al. 2018).
To make the datasets more comparable (Figure 1), we filtered out samples from spatially distant localities between marker datasets.

## 2.3 Population structure
We conducted model-based genetic structure analyses in STRUCTURE v2.3.4 (Pritchard et al. 2000). For each dataset, we performed 10 replicate runs from K = 1 to K = 8, since the best number of clusters fall within this range for both species (Gutiérrez-Rodríguez et al. 2017; Sánchez-Montes et al. 2019). We used an admixture model with correlated allele frequencies (Falush et al. 2003) and no prior information of sample origin. For the microtellite data, we used the same run lengths as in their original publications: 5*10^5 burnin followed by 1*10^6 iterations. For the SNP datasets, the run lengths were shorter, a burnin of 3*10^4 steps followed by 1 * 10^4 iterations, as for SNP data chain often converge fast (Pritchard, google groups...?). For the SNP data, we estimated lambda for K = 1, by averaging lambda estimates across 3 replicate runs. The values of lambda (0.60 for *H. molleri* and 0.69 for *P. cultripes*) were then used across all runs for the SNP data, whereas for microsatellites lambda was fixed to 1. These lower values of lambda can improve the modeling of correlated allele frequencies when using SNPs, where often the data is skewed toward rare alleles (Falush et al. 2003). STRUCTURE was run in parallel in 8 cores using Structure_threader (Pina-Martins et al. 2017), recording steps to log files every 50 and 5000 iterations for the SNP and microsatellite data, respectively. We run STRUCTURE analysis with random subsets of loci (n = 200, 500, 1000, 3000, 5000, 10000 and 20000) from the original matrix of SNPs for *P. cultripes* and the same STRUCTURE parameters.
The convergence between the 10 replicate runs for each K was evaluated in R using Gelman and Rubin's convergence diagnostic (GR diagnostic; Gelman and Rubin 1992), with function *gelman.diag* in package *coda* (Plummer et al. 2006).

# 3. RESULTS

After filtering (Figure S1), the dataset kept for analysis included a total `r totalind` individuals from `r nrow(samples)` localities: `r indH` individuals of *Hyla molleri* (microtellites: n = `r indHm`, `r locHm` loci, `r missHm` % missing data; SNPs: n = `r indHs`, `r locHs` loci, `r missHs` % missing data), from `r popH` localities, and `r indP` individuals of *Pelobates cultripes* (microtellites: n = `r indPm`, `r locPm` loci, `r missPm` % missing data; SNPs: n = `r indPs`, `r locPs` loci, `r missPs` % missing data), from `r popP` localities (Figure 1; Table S1).

## 3.1 Model-based clustering
The convergence of the chains as measured by GR diagnostic varied between speciesa and markers. Often, values below 1.05 indicate good convergence (Vats and Knudson 2018 Jeff should check this ref). For SNPs, the chains from each K in STRUCTURE converged well up to K = 2 in *H. molleri* and up to K = 6 in the case of *P. cultripes*. Reducing the number of loci did not affect the pattern of convergence across K in a consistent manner (Figures S2.1, S2.2, S2.5 - S2.11; Figure S3.1).
For microsatellite data, the chains converged consistently up to K = 6 in both species, although the GR diagnostic slightly peaked for K = 5 in *P. cultripes* (Figures S2.3 - S2.4; Figure S3.1).
